### 2015-10-17 doc/paper.tex
Start working on the paper

\documentclass[11pt]{amsart}

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath, stmaryrd, scalerel}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{proof}
\usepackage{supertabular}
\usepackage{unicode-math}
\usepackage{xltxtra}
\usepackage{xunicode}

\usepackage{setspace,url,hyperref}

% fonts
\setromanfont[Mapping=tex-text, Ligatures=Rare, Numbers=OldStyle]{Linux Libertine O}
\setmathfont{Asana Math}

\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}
\newtheorem{example}[thm]{Example}
\theoremstyle{remark}
\newtheorem{remark}[thm]{Remark}
\numberwithin{equation}{section}

\lstset{
  language=[Objective]Caml,
  extendedchars=\true,
  inputencoding=utf8,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  basicstyle=\ttfamily,
  xleftmargin=6mm,
  framesep=4mm,
  moredelim=*[s][\itshape]{(*}{*)},
  moredelim=[is][\textcolor{darkgray}]{§}{§},
  escapechar=°,
  keywords=[6]{signature, structure, struct type, sig, end},
  keywordstyle=\color{MidnightBlue},
  stringstyle=\color[rgb]{0.545098, 0.278431, 0.364706},
  commentstyle=\color[rgb]{0.698039, 0.133333, 0.133333},
  numberstyle=\color[rgb]{0.372549, 0.619608, 0.627451},
  literate=
    {->}{{$\rightharpoonup$}}{1}
    {*}{{$\otimes$}}{1}
}

\def\InputModeColorName{MidnightBlue}
\def\OutputModeColorName{Maroon}
\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}
\newcommand\Quote[1]{{\color{Gray}{«}}#1{\color{Gray}{»}}}

\newcommand\HypJ[2]{#1\ \ (#2)}
\newcommand\GenJ[2]{\vert_{\InputMode{#1}}\; #2}
\newcommand\JJ{J}
\newcommand\Tuple[1]{\left\langle#1\right\rangle}
\newcommand\HTuple[1]{\left\lBrace#1\right\rBrace}
\newcommand\RSet{\mathfrak{R}}
\newcommand\JSet{\mathfrak{J}}

\newcommand\IsTac[1]{\InputMode{#1}\ \textit{tactic}}

%% subgoals
\newcommand\SG[2]{\InputMode{#1}\,{\color{Black}:}\,\OutputMode{#2}}

%% refinement
\newcommand\Refine[4]{\InputMode{#1}\Vdash\InputMode{#2}\Rightarrow\OutputMode{#4}\dashv\OutputMode{#3}}
\newcommand\NoRefine[2]{\InputMode{#1}\Vdash\InputMode{#2}\Uparrow}

\newcommand\Member[2]{\InputMode{#1}\in\InputMode{#2}}
\newcommand\IdTac{\mathtt{id}}
\newcommand\FailTac{\mathtt{fail}}
\newcommand\OrElseTac[2]{#1\vert#2}
\newcommand\ThenTac[2]{#1;#2}
\newcommand\ThenLTac[2]{#1;[#2]}
\newcommand\ProdIntroRule{\mathtt{ProdIntro}}
\newcommand\DProdIntroRule[1]{\mathtt{DepProdIntro}\{#1\}}
\newcommand\DProdIntroRuleNullary{\mathtt{DepProdIntro}}
\newcommand\TyProd[2]{#1 \times #2}
\newcommand\TyDProd[3]{(#2:#1) \times #3}
\newcommand\TyMember[2]{#1\in#2}
\newcommand\Dom[1]{\left\vert#1\right\vert}
\newcommand\SubsetEq[2]{\InputMode{#1}\subseteq\InputMode{#2}}
\newcommand\FV[1]{\textit{FV}\left(#1\right)}
\newcommand\Subst[3]{\left[#1\;/\;#2\right]#3}

\begin{document}

\title{Dependency in Refinement Logics}

\author{Jonathan Sterling}
\address{}
\maketitle

\onehalfspacing

Over the past four decades, there have been many incarnations of the \emph{LCF}
(Logic for Computable Functions) interface, but the one that will concern us
here is its extension to \emph{LCF with Validations} as found in Cambridge LCF,
which outfits the proving activity with the synthesis of explicit evidence.
The Cambridge LCF signature is as follows:

%% cite "Logic and Computation: Interactive Proof with Cambridge LCF"

\begin{lstlisting}
  signature CAMBRIDGE_LCF =
  sig
    type form
    type thm
    type proof = thm list -> thm
    type goal = form list * form
    type tactic = goal -> goal list * proof
  end
\end{lstlisting}

The \verb!goal! type represents sequents, where \verb!form! represents logical
propositions. The LCF methodology, however, may be used to give a refinement
treatment to many kinds of judgments, not just sequents; therefore, it
is better to not include this in the core signature at all, and make the
type of judgments abstract; for clarity, we will replace the name \verb!thm!
with \verb!evidence!.

\begin{lstlisting}
  signature LCF =
  sig
    type judgment
    type evidence
    type proof = evidence list -> evidence
    type tactic = judgment -> judgment list * proof
  end
\end{lstlisting}

Then, a tactic is something that refines a judgment to a list of judgments (its
subgoals), and synthesizes its evidence (provided the evidence of its
subgoals). There are many different tactics which can be implemented
generically over this signature; here are a few:

\begin{lstlisting}
  signature TACTICALS =
  sig
    structure Lcf : LCF
    val ID : Lcf.tactic
    val FAIL : Lcftactic
    val THEN : Lcf.tactic * Lcf.tactic -> Lcf.tactic
    val THENL : Lcf.tactic * Lcf.tactic list -> Lcf.tactic
  end
\end{lstlisting}

\section{modernized lcf: the logic of tactics}

In order to study the design space for LCF refiners, we would like to give a
judgmental characterization of tactic systems. First, we will recognize that
the type of validations (\verb!proof!) in the ML implementation is essentially
a HOAS (higher-order abstract syntax) encoding of a hypothetical proof or
synthesis. Then, it is clear that the list of subgoals generated by a tactic
are a context which explains the variables bound by the hypothetical synthesis.

To make the preceding observations precise, we can recover the character of an
LCF refiner in a single form of judgment,
\framebox{$\Refine{\JJ}{\tau}{\Psi}{E}$}, where $\JJ$ is a judgment of the
logical theory, $\tau$ is a tactic, $\Psi$ is a context of judgments,
representing the subgoals generated by the tactic $\tau$, and $E$ is the
synthesis of the judgment $\JJ$, binding variables $\Dom{\Psi}$ which represent
the syntheses of the subgoals. The meaning of this assertion is that $\tau$ is
applicable to demonstrating the judgment $\JJ$, producing synthesis $e$ under
the assumptions that the judgments in $\Psi$ can be demonstrated. In addition
to the refinement judgment, there is also refinement failure
$\NoRefine{\JJ}{\tau}$, which expresses the inapplicability of $\tau$ to $\JJ$.
We require as a matter of policy that the applicability of a tactic be
discrete, and in practice, we leave implicit the rules which explain the
inapplicability of a tactic.

\begin{remark}
  The refinement judgments $\Refine{\JJ}{\tau}{\Psi}{E}$ and
  $\NoRefine{\JJ}{\tau}$ are not higher-order judgments, because the variable
  $\JJ$ is ranges not over judgments of the refinement theory, but of the
  object theory.
\end{remark}

Let $\JSet$ be the open-ended collection of judgments in our logical theory,
and let $\RSet$ be a collection of rule names. Each rule $\Member{R}{\RSet}$
must be interpretable as a tactic, i.e.\ the meaning of the assertions
$\Refine{\JJ}{R}{\Psi}{E}$ and $\NoRefine{\JJ}{R}$ must be explained for
$\Member{J}{\JSet}$. We say $\IsTac{\tau}$ in case for all object-judgments
$\Member{\JJ}{\JSet}$, the assertion conditions for
$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$ are disjoint, and
moreover, if $\Refine{\JJ}{\tau}{\Psi}{E}$, then
$\SubsetEq{\FV{E}}{\Dom{\Psi}}$.

In practice, we will explain only the assertion conditions for one of
$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$, and take the other to
be its complement.

Numerous general purpose tactics can be defined over a logical theory,
including identity, failure, disjunction and sequencing:

\begin{gather*}
  \infer{
    \Refine{\JJ}{\IdTac}{\SG{\alpha}{\JJ}}{\alpha}
  }{
  }
  \qquad
  \infer{
    \NoRefine{\JJ}{\FailTac}
  }{
  }\\[6pt]
  \infer{
    \Refine{\JJ}{\OrElseTac{\tau_1}{\tau_2}}{\Psi}{E}
  }{
    \Refine{\JJ}{\tau_1}{\Psi}{E}
  }
  \qquad
  \infer{
    \Refine{\JJ}{\OrElseTac{\tau_1}{\tau_2}}{\Psi}{E}
  }{
    \NoRefine{\JJ}{\tau_1} &
    \Refine{\JJ}{\tau_2}{\Psi}{E}
  }\\[6pt]
  \infer{
    \Refine{\JJ}{\ThenTac{\tau_1}{\tau_2}}{\bigoplus_{\Dom\Phi}\Psi}{\Subst{F}{\Dom\Phi}{E}}
  }{
    \Refine{\JJ}{\tau_1}{\Phi}{E} &
    \GenJ{\alpha}{
      \HypJ{
        \Refine{\Phi(\alpha)}{\tau_2}{\Psi_\alpha}{F_\alpha}
      }{
        \Member{\alpha}{\Dom\Phi}
      }
    }
  }
\end{gather*}

\begin{remark}
  The nominal treatment that we have given here allows for a much more
  economical presentation of the standard tacticals, which are quite ardruous
  to define in the HOAS treatment used in ML implementations.
\end{remark}

\begin{thm}
  The above rules all define valid tactics:.
  \begin{enumerate}
    \item $\IsTac{\IdTac}$.
    \item $\IsTac{\FailTac}$.
    \item If $\IsTac{\tau_1}$ and $\IsTac{\tau_2}$, then
      $\IsTac{\OrElseTac{\tau_1}{\tau_2}}$.
    \item If $\IsTac{\tau_1}$ and $\IsTac{\tau_2}$, then
      $\IsTac{\ThenTac{\tau_1}{\tau_2}}$.
  \end{enumerate}
\end{thm}
\begin{proof}
  It suffices to verify that the synthesis $E$ of each tactical is well-scoped
  in $\Psi$; in each case, this follows by induction.
  \begin{enumerate}
    \item[(1--2)] Immediate.
    \item[(3)]
      By induction on derivations of
      $\Refine{\JJ}{\OrElseTac{\tau_1}{\tau_1}}{E}{\Psi}$.
      \begin{description}
        \item[Case $\Refine{\JJ}{\tau_1}{\Psi}{E}$] Validity follows from the
          inductive hypothesis $\IsTac{\tau_1}$.
        \item[Case $\NoRefine{\JJ}{\tau_1}$] Validity follows from the
          inductive hypothesis $\IsTac{\tau_2}$.
      \end{description}

    \item[(4)] We need to show that
      $\SubsetEq{\FV{\Subst{F}{\Dom\Phi}{E}}}{\Dom{\bigoplus_{\Dom\Phi}\Psi}}$.
      From our inductive hypotheses, it is evident that
      $\SubsetEq{\FV{E}}{\Dom\Phi}$ and that for each $\Member{\alpha}{\Phi}$,
      $\SubsetEq{\FV{F_\alpha}}{\Dom{\Psi_\alpha}}$; all of the free variables
      of the term got by substituting each $F_\alpha$ for
      $\Member{\alpha}{\Dom\Phi}$ clearly reside in one of the fibres of the
      family of contexts $\Psi$, and so they must comprise a subset of the
      union $\bigoplus_{\Dom\Phi}\Psi$ of $\Psi$'s fibres.

  \end{enumerate}
\end{proof}

\end{document}

### 2015-10-17 doc/paper.tex
typo

@@
     val ID : Lcf.tactic
-    val FAIL : Lcftactic
+    val FAIL : Lcf.tactic
     val THEN : Lcf.tactic * Lcf.tactic -> Lcf.tactic
@@
 In practice, we will explain only the assertion conditions for one of
-$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$, and take the other to
-be its complement.
+$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$, and implicitly take
+the other to be its complement.
 

### 2015-10-17 doc/paper.tex
add short abstract
typo

@@
 \newcommand\Subst[3]{\left[#1\;/\;#2\right]#3}
+\newcommand\DepLCF{\textbf{\emph{Dependent LCF}}}
 
@@
 \maketitle
+
+\begin{abstract}
+
+  We propose a new generalization of LCF-style refinement logics called \DepLCF,
+  which allows the definition of refinement rules which express a dependency
+  between subgoals, without the use of unification variables.
+
+\end{abstract}
 
@@
   $\NoRefine{\JJ}{\tau}$ are not higher-order judgments, because the variable
-  $\JJ$ is ranges not over judgments of the refinement theory, but of the
+  $\JJ$ ranges not over judgments of the refinement theory, but of the
   object theory.

### 2015-10-17 doc/paper.tex
Cleanup

@@
 \newcommand\Subst[3]{\left[#1\;/\;#2\right]#3}
+\newcommand\ModLCF{\textbf{\emph{Modernized LCF}}}
 \newcommand\DepLCF{\textbf{\emph{Dependent LCF}}}
@@
 In order to study the design space for LCF refiners, we would like to give a
-judgmental characterization of tactic systems. First, we will recognize that
-the type of validations (\verb!proof!) in the ML implementation is essentially
-a HOAS (higher-order abstract syntax) encoding of a hypothetical proof or
-synthesis. Then, it is clear that the list of subgoals generated by a tactic
-are a context which explains the variables bound by the hypothetical synthesis.
+judgmental characterization of tactic systems, which we will call $\ModLCF$.
+To begin with, note that the type of validations (\verb!proof!) in the ML
+implementation is is essentially a HOAS (higher-order abstract syntax) encoding
+of a hypothetical proof or synthesis of a judgment.  With this insight in hand,
+we are in a position to unify the list of subgoals and the validation generated
+by a tactic into a single concept, namely that of a \emph{hypothetical proof}
+$E$ whose free variables are explained in a context $\Psi$ of subgoals.
 
-To make the preceding observations precise, we can recover the character of an
-LCF refiner in a single form of judgment,
-\framebox{$\Refine{\JJ}{\tau}{\Psi}{E}$}, where $\JJ$ is a judgment of the
+To make the preceding observations precise, we can characterize the behavior of
+a \ModLCF\ refiner judgmentally via two forms of judgment,
+\framebox{$\Refine{\JJ}{\tau}{\Psi}{E}$} and \framebox{$\NoRefine{\JJ}{\tau}$}, where $\JJ$ is a judgment of the
 logical theory, $\tau$ is a tactic, $\Psi$ is a context of judgments,
@@
 synthesis of the judgment $\JJ$, binding variables $\Dom{\Psi}$ which represent
-the syntheses of the subgoals. The meaning of this assertion is that $\tau$ is
-applicable to demonstrating the judgment $\JJ$, producing synthesis $e$ under
-the assumptions that the judgments in $\Psi$ can be demonstrated. In addition
-to the refinement judgment, there is also refinement failure
-$\NoRefine{\JJ}{\tau}$, which expresses the inapplicability of $\tau$ to $\JJ$.
-We require as a matter of policy that the applicability of a tactic be
-discrete, and in practice, we leave implicit the rules which explain the
-inapplicability of a tactic.
+the syntheses of the subgoals.
+
+The meaning of $\Refine{\JJ}{\tau}{\Psi}{E}$ is that $\tau$ is applicable to
+demonstrating the judgment $\JJ$, producing synthesis $e$ under the assumptions
+that the judgments in $\Psi$ can be demonstrated. The divergence judgment
+$\NoRefine{\JJ}{\tau}$ expresses the inapplicability of $\tau$ to $\JJ$.  In
+practice, we will explain only the assertion conditions for one of
+$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$, and implicitly take
+the other to be its complement.
 
@@
 $\SubsetEq{\FV{E}}{\Dom{\Psi}}$.
-
-In practice, we will explain only the assertion conditions for one of
-$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$, and implicitly take
-the other to be its complement.
 

### 2015-10-17 doc/paper.tex
Add constructible subgoals section

@@
-\documentclass[11pt]{amsart}
+\documentclass[11pt]{article}
 
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\usepackage{amsmath, stmaryrd, scalerel}
+\usepackage{amsmath, amsthm, stmaryrd, scalerel}
 \usepackage{listings}
@@
 % fonts
-\setromanfont[Mapping=tex-text, Ligatures=Rare, Numbers=OldStyle]{Linux Libertine O}
+% \setromanfont[Mapping=tex-text, Ligatures=Rare, Numbers=OldStyle]{Linux Libertine O}
 \setmathfont{Asana Math}
@@
 \newcommand\Member[2]{\InputMode{#1}\in\InputMode{#2}}
+\newcommand\MemberUnmoded[2]{#1\in #2}
+\newcommand\IsTrue[1]{\InputMode{#1}\ \textit{true}}
+\newcommand\IsTrueUnmoded[1]{#1\ \textit{true}}
 \newcommand\IdTac{\mathtt{id}}
@@
 \title{Dependency in Refinement Logics}
-
+\date{}
 \author{Jonathan Sterling}
-\address{}
 \maketitle
@@
 \end{abstract}
-
-\onehalfspacing
 
@@
 
-\section{modernized lcf: the logic of tactics}
+\section{\ModLCF: the logic of tactics}
 
@@
 
+\section{\ModLCF\ and the constructible subgoals property}
+
+An LCF-style refiner has a property called \emph{constructible subgoals}, which
+means that the subgoals incurred by a rule or tactic may be constructed
+independently, using only the statement of the main goal. One unfortunate
+consequence of this principle is that it is impossible to define a refinement
+rule which expresses a dependency between subgoals. The canonical example is
+the introduction rule for dependent pairs in Type Theory:
+\[
+  \infer{
+    \Member{\Tuple{M,N}}{\TyDProd{A}{x}{B}}
+  }{
+    \Member{M}{A} &
+    \Member{N}{\Subst{M}{x}{B}}
+  }
+\]
+
+However, consider what a refinement rule for this would look like; we would
+like to take the goal $\IsTrue{\TyDProd{A}{x}{B}}$ to two subgoals (one for
+each conjunct); the first subgoal $\IsTrue{A}$ is clear enough, but it is not
+possible to even write down the second subgoal until we know the synthesis of
+the first one.
+
+Because a tactic in \ModLCF\ produces only a \emph{context} of independent
+subgoals, we cannot give a refinement treatment to this subgoal, and must
+instead write a family of refinement rules $\DProdIntroRule{w}$ fibred over
+witnesses of the left conjunct $w$:
+
+\[
+  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRule{w}}{%
+    \left\{\begin{array}{l}
+      \SG{\alpha}{\MemberUnmoded{w}{A}}\\
+      \SG{\beta}{\IsTrueUnmoded{\Subst{w}{x}{B}}}
+    \end{array}\right.
+  }{\Tuple{w,\beta}}
+\]
+
+This is clearly unsatisfactory, since it breaks the natural flow of proof
+development, whereby multiple goals may be refined simultaneously without
+committing in advance to a particular solution. However, a more palatable
+refinement rule that allows $\IsTrue{A}$ to be demonstrated by refinement is
+simply not expressible in \ModLCF, since the sense of the second subgoal cannot
+be expressed except by referring to the synthesis of the first subgoal.
+
 \end{document}

### 2015-10-17 doc/paper.tex
introduce dependent lcf

@@
 \newcommand\ThenLTac[2]{#1;[#2]}
-\newcommand\ProdIntroRule{\mathtt{ProdIntro}}
-\newcommand\DProdIntroRule[1]{\mathtt{DepProdIntro}\{#1\}}
-\newcommand\DProdIntroRuleNullary{\mathtt{DepProdIntro}}
+\newcommand\DProdIntroRule[1]{\mathtt{PairIntro}\{#1\}}
+\newcommand\DProdIntroRuleNullary{\mathtt{PairIntro}}
 \newcommand\TyProd[2]{#1 \times #2}
@@
 \[
-  \infer{
+  \infer[\textit{PairIntro}]{
     \Member{\Tuple{M,N}}{\TyDProd{A}{x}{B}}
@@
 
+\section{\DepLCF\ and generalized refinement rules}
+
+At the crux of our problem is that a tactic produces a context of subgoals
+without any dependencies; if we were to construe the judgment
+$\Refine{\JJ}{\tau}{\Psi}{E}$ as synthesizing a \emph{telescope} $\Psi$ rather
+than a mere context, a proper refinement rule for \emph{PairIntro} would be
+within reach. In fact, whilst the ML signature for LCF refiners rules out this
+interpretation, the notation we have used for \ModLCF\ immediately suggests
+this generalization. Going forward, we will call the theory \DepLCF\ when we
+take $\Psi$ to be a telescope rather than a context.
+
+We can now encode \emph{PairIntro} as a single refinement rule,
+$\DProdIntroRuleNullary$:
+
+\[
+  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRuleNullary}{%
+    \left\{\begin{array}{l}
+        \SG{\alpha}{\IsTrueUnmoded{A}}\\
+        \SG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
+    \end{array}\right.
+  }{
+    \Tuple{\alpha,\beta}
+  }
+\]
+
+All that remains is to give a new definition of the sequencing tactical
+$\ThenTac{\tau_1}{\tau_2}$ which accounts for this dependency.
+
 \end{document}

### 2015-10-17 doc/paper.tex
Implement THEN in the paper

@@
 \usepackage{amsmath, amsthm, stmaryrd, scalerel}
+\usepackage{mathtools}
 \usepackage{listings}
@@
 %% subgoals
-\newcommand\SG[2]{\InputMode{#1}\,{\color{Black}:}\,\OutputMode{#2}}
+\newcommand\ISG[2]{\InputMode{#1}\,{\color{Black}:}\,\InputMode{#2}}
+\newcommand\OSG[2]{\InputMode{#1}\,{\color{Black}:}\,\OutputMode{#2}}
 
@@
 \newcommand\NoRefine[2]{\InputMode{#1}\Vdash\InputMode{#2}\Uparrow}
+
+\newcommand\ThenStateTrans[5]{%
+  \InputMode{#2}\dashv\InputMode{#3}%
+  \hookrightarrow_\InputMode{#1}%
+  \OutputMode{#4}\dashv\OutputMode{#5}%
+}
 
@@
   \infer{
-    \Refine{\JJ}{\IdTac}{\SG{\alpha}{\JJ}}{\alpha}
+    \Refine{\JJ}{\IdTac}{\OSG{\alpha}{\JJ}}{\alpha}
   }{
@@
     \left\{\begin{array}{l}
-      \SG{\alpha}{\MemberUnmoded{w}{A}}\\
-      \SG{\beta}{\IsTrueUnmoded{\Subst{w}{x}{B}}}
+      \OSG{\alpha}{\MemberUnmoded{w}{A}}\\
+      \OSG{\beta}{\IsTrueUnmoded{\Subst{w}{x}{B}}}
     \end{array}\right.
@@
 
-We can now encode \emph{PairIntro} as a single refinement rule,
+We can now encode \emph{PairIntro} with a single refinement rule,
 $\DProdIntroRuleNullary$:
@@
     \left\{\begin{array}{l}
-        \SG{\alpha}{\IsTrueUnmoded{A}}\\
-        \SG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
+        \OSG{\alpha}{\IsTrueUnmoded{A}}\\
+        \OSG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
     \end{array}\right.
@@
 All that remains is to give a new definition of the sequencing tactical
-$\ThenTac{\tau_1}{\tau_2}$ which accounts for this dependency.
+$\ThenTac{\tau_1}{\tau_2}$ which accounts for this dependency. In order to do
+this, we will need to implement an auxiliary judgment
+\framebox{$\ThenStateTrans{\tau}{E}{\Phi}{E'}{\Phi'}$} that allows us to
+gradually apply the tactic $\tau$ to a proof state, propagating refinements
+rightward through substitution. This judgment is defined by recursion on the
+telescope $\Phi$, viewed as a \emph{cons}-list:
+
+\begin{gather*}
+  \infer{
+    \ThenStateTrans{\tau}{E}{\cdot}{E}{\cdot}
+  }{
+  }\\[6pt]
+  \infer{
+    \ThenStateTrans{\tau}{E}{\ISG{\alpha}{\JJ},\Phi}{E'}{\Psi_\alpha\oplus\Phi'}
+  }{
+    \Refine{\JJ}{\tau}{\Psi_\alpha}{E_\alpha} &
+    \ThenStateTrans{\tau}{%
+      \Subst{E_\alpha}{\alpha}{E}%
+    }{
+      \Subst{E_\alpha}{\alpha}{\Phi}
+    }{E'}{\Phi'}
+  }
+\end{gather*}
+
+Now, the sequencing tactical is readily definable:
+\[
+  \infer{
+    \Refine{\JJ}{\ThenTac{\tau_1}{\tau_2}}{\Phi'}{E'}
+  }{
+    \Refine{\JJ}{\tau_1}{\Phi}{E} &
+    \ThenStateTrans{\tau_2}{E}{\Phi}{E'}{\Phi'}
+  }
+\]
 

### 2015-10-17 doc/paper.tex
typo

@@
 Because a tactic in \ModLCF\ produces only a \emph{context} of independent
-subgoals, we cannot give a refinement treatment to this subgoal, and must
+subgoals, we cannot give a refinement treatment to this rule, and must
 instead write a family of refinement rules $\DProdIntroRule{w}$ fibred over

### 2015-10-17 doc/paper.tex
unify some terminology

@@
 type of judgments abstract; for clarity, we will replace the name \verb!thm!
-with \verb!evidence!.
+with \verb!synthesis!.
 
@@
     type judgment
-    type evidence
-    type proof = evidence list -> evidence
+    type synthesis
+    type proof = synthesis list -> synthesis
     type tactic = judgment -> judgment list * proof
@@
 Then, a tactic is something that refines a judgment to a list of judgments (its
-subgoals), and synthesizes its evidence (provided the evidence of its
+subgoals), and synthesizes its evidence (provided the syntheses of its
 subgoals). There are many different tactics which can be implemented
@@
 
-At the crux of our problem is that a tactic produces a context of subgoals
+At the crux of our problem is the fact that a tactic produces a context of subgoals
 without any dependencies; if we were to construe the judgment

### 2015-10-17 doc/paper.tex
add a forgotten \framebox

@@
 $\Refine{\JJ}{R}{\Psi}{E}$ and $\NoRefine{\JJ}{R}$ must be explained for
-$\Member{J}{\JSet}$. We say $\IsTac{\tau}$ in case for all object-judgments
+$\Member{J}{\JSet}$. We say \framebox{$\IsTac{\tau}$} in case for all object-judgments
 $\Member{\JJ}{\JSet}$, the assertion conditions for

### 2015-10-17 doc/paper.tex
Add some dots

@@
       \begin{description}
-        \item[Case $\Refine{\JJ}{\tau_1}{\Psi}{E}$] Validity follows from the
+        \item[Case $\Refine{\JJ}{\tau_1}{\Psi}{E}$.] Validity follows from the
           inductive hypothesis $\IsTac{\tau_1}$.
-        \item[Case $\NoRefine{\JJ}{\tau_1}$] Validity follows from the
+        \item[Case $\NoRefine{\JJ}{\tau_1}$.] Validity follows from the
           inductive hypothesis $\IsTac{\tau_2}$.

### 2015-10-17 doc/paper.tex
notation switch; wordsmithing

@@
   \InputMode{#2}\dashv\InputMode{#3}%
-  \hookrightarrow_\InputMode{#1}%
+  \Downarrow_\InputMode{#1}%
   \OutputMode{#4}\dashv\OutputMode{#5}%
@@
 committing in advance to a particular solution. However, a more palatable
-refinement rule that allows $\IsTrue{A}$ to be demonstrated by refinement is
+rule that allows $\IsTrue{A}$ to be demonstrated by refinement is
 simply not expressible in \ModLCF, since the sense of the second subgoal cannot

### 2015-10-18 doc/paper.tex
typo

@@
 To begin with, note that the type of validations (\verb!proof!) in the ML
-implementation is is essentially a HOAS (higher-order abstract syntax) encoding
+implementation is essentially a HOAS (higher-order abstract syntax) encoding
 of a hypothetical proof or synthesis of a judgment.  With this insight in hand,
@@
 
+\section{ML Implementation of \DepLCF}
+
 \end{document}

### 2015-10-18 doc/paper.tex
Generalize to THENL tactical, using families of tactics

@@
 
+\newcommand\MkSet[1]{\left\{#1\right\}}
+\newcommand\MkFam[3]{\MkSet{#1_{#2}}_{#2\in #3}}
+
 %% subgoals
@@
 
-\newcommand\ThenStateTrans[5]{%
+\newcommand\ThenMultiAux[5]{%
   \InputMode{#2}\dashv\InputMode{#3}%
-  \Downarrow_\InputMode{#1}%
+  \Vdash^\star\InputMode{#1}\Rightarrow%
   \OutputMode{#4}\dashv\OutputMode{#5}%
@@
 \newcommand\DepLCF{\textbf{\emph{Dependent LCF}}}
+
+\newcommand\IsEquiv[2]{\InputMode{#1}\equiv\InputMode{#2}}
 
@@
   \infer{
-    \Refine{\JJ}{\ThenTac{\tau_1}{\tau_2}}{\bigoplus_{\Dom\Phi}\Psi}{\Subst{F}{\Dom\Phi}{E}}
+    \Refine{\JJ}{\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{\Dom\Phi}}}{\bigoplus_{\Dom\Phi}\mathbf\Psi}{\Subst{F}{\Dom\Phi}{E}}
   }{
@@
       \HypJ{
-        \Refine{\Phi(\alpha)}{\tau_2}{\Psi_\alpha}{F_\alpha}
+        \Refine{\Phi(\alpha)}{\tau_\alpha}{\Psi_\alpha}{F_\alpha}
       }{
@@
 
+The sequencing tactical
+$\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}$ corresponds to
+\verb!THENL! in LCF; the uniform sequencing tactical \verb!THEN! may be
+recovered as a definitional extension, with its second argument given as the
+constant family:
+\[
+    \InputMode{\ThenTac{\tau_1}{\tau_2}}
+    \triangleq
+    \OutputMode{\ThenTac{\tau_1}{\MkSet{\tau_2}}}
+\]
+
 \begin{thm}
@@
       $\IsTac{\OrElseTac{\tau_1}{\tau_2}}$.
-    \item If $\IsTac{\tau_1}$ and $\IsTac{\tau_2}$, then
-      $\IsTac{\ThenTac{\tau_1}{\tau_2}}$.
+
+    \item If $\IsTac{\tau_1}$ and for any $\Member{\alpha}{I}$,
+      $\IsTac{\tau_\alpha}$, then
+      $\IsTac{\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}}$.
+
+     %% If $\IsTac{\tau_1}$ and $\IsTac{\tau_2}$, then
+     %% $\IsTac{\ThenTac{\tau_1}{\tau_2}}$.
   \end{enumerate}
@@
 
-    \item[(4)] We need to show that
+    \item[(4)] It suffices to verify the cases where
+      $\IsEquiv{\Dom\Phi}{I}$, since otherwise
+      $\NoRefine{\JJ}{\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}}$.
+      We need to show that
       $\SubsetEq{\FV{\Subst{F}{\Dom\Phi}{E}}}{\Dom{\bigoplus_{\Dom\Phi}\Psi}}$.
@@
 All that remains is to give a new definition of the sequencing tactical
-$\ThenTac{\tau_1}{\tau_2}$ which accounts for this dependency. In order to do
-this, we will need to implement an auxiliary judgment
-\framebox{$\ThenStateTrans{\tau}{E}{\Phi}{E'}{\Phi'}$} that allows us to
-gradually apply the tactic $\tau$ to a proof state, propagating refinements
-rightward through substitution. This judgment is defined by recursion on the
-telescope $\Phi$, viewed as a \emph{cons}-list:
+$\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}$ which accounts for this dependency.
+In order to do this, we will need to implement an auxiliary judgment
+\framebox{$\ThenMultiAux{\MkFam{\tau}{\alpha}{I}}{E}{\Phi}{E'}{\Phi'}$} that
+allows us to apply the family of tactics $\MkFam{\tau}{\alpha}{I}$ pointwise to
+a proof state, simultaneously propagating refinements rightward through
+substitution. This judgment, which presupposes $\IsEquiv{I}{\Dom\Phi}$, is defined
+by recursion on the telescope $\Phi$, viewed as a \emph{cons}-list:
 
@@
   \infer{
-    \ThenStateTrans{\tau}{E}{\cdot}{E}{\cdot}
+    \ThenMultiAux{\MkSet{}}{E}{\cdot}{E}{\cdot}
   }{
   }\\[6pt]
+  %
   \infer{
-    \ThenStateTrans{\tau}{E}{\ISG{\alpha}{\JJ},\Phi}{E'}{\Psi_\alpha\oplus\Phi'}
+    \ThenMultiAux{
+      \MkFam{\tau}{\beta}{\Dom{\ISG{\alpha}{\JJ},\Phi}}
+    }{E}{
+      \ISG{\alpha}{\JJ},\Phi
+    }{E'}{
+      \Psi_\alpha\oplus\Phi'
+    }
   }{
     \Refine{\JJ}{\tau}{\Psi_\alpha}{E_\alpha} &
-    \ThenStateTrans{\tau}{%
+    \ThenMultiAux{
+      \MkFam{\tau}{\beta}{\Phi}
+    }{
       \Subst{E_\alpha}{\alpha}{E}%
@@
 
-Now, the sequencing tactical is readily definable:
+Now, the sequencing tactical is readily definable, presupposing $\IsEquiv{I}{\Dom\Phi}$:
 \[
   \infer{
-    \Refine{\JJ}{\ThenTac{\tau_1}{\tau_2}}{\Phi'}{E'}
+    \Refine{\JJ}{\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}}{\Phi'}{E'}
   }{
     \Refine{\JJ}{\tau_1}{\Phi}{E} &
-    \ThenStateTrans{\tau_2}{E}{\Phi}{E'}{\Phi'}
+    \ThenMultiAux{
+      \MkFam{\tau}{\alpha}{I}
+    }{E}{\Phi}{E'}{\Phi'}
   }
@@
 
-\section{ML Implementation of \DepLCF}
+%% \section{ML Implementation of \DepLCF}
 

### 2015-10-18 doc/paper.tex
Fix missing subscript
demonstrate rules for dependent pair, function

@@
 \usepackage{setspace,url,hyperref}
+
+\definecolor{dark-gray}{gray}{0.20}
 
@@
 
-\def\InputModeColorName{MidnightBlue}
-\def\OutputModeColorName{Maroon}
-\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
-\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}
+\def\IModeColorName{MidnightBlue}
+\def\OModeColorName{Maroon}
+\newcommand\IMode[1]{{\color{\IModeColorName}{#1}}}
+\newcommand\OMode[1]{{\color{\OModeColorName}{#1}}}
 \newcommand\Quote[1]{{\color{Gray}{«}}#1{\color{Gray}{»}}}
@@
 \newcommand\HypJ[2]{#1\ \ (#2)}
-\newcommand\GenJ[2]{\vert_{\InputMode{#1}}\; #2}
+\newcommand\GenJ[2]{\vert_{\IMode{#1}}\; #2}
 \newcommand\JJ{J}
-\newcommand\Tuple[1]{\left\langle#1\right\rangle}
-\newcommand\HTuple[1]{\left\lBrace#1\right\rBrace}
 \newcommand\RSet{\mathfrak{R}}
@@
 
-\newcommand\IsTac[1]{\InputMode{#1}\ \textit{tactic}}
+\newcommand\IsTac[1]{\IMode{#1}\ \textit{tactic}}
 
@@
 %% subgoals
-\newcommand\ISG[2]{\InputMode{#1}\,{\color{Black}:}\,\InputMode{#2}}
-\newcommand\OSG[2]{\InputMode{#1}\,{\color{Black}:}\,\OutputMode{#2}}
+\newcommand\ISG[2]{\IMode{#1}\,{\color{Black}:}\,\IMode{#2}}
+\newcommand\OSG[2]{\IMode{#1}\,{\color{Black}:}\,\OMode{#2}}
 
 %% refinement
-\newcommand\Refine[4]{\InputMode{#1}\Vdash\InputMode{#2}\Rightarrow\OutputMode{#4}\dashv\OutputMode{#3}}
-\newcommand\NoRefine[2]{\InputMode{#1}\Vdash\InputMode{#2}\Uparrow}
+\newcommand\Refine[4]{\IMode{#1}\Vdash\IMode{#2}\Rightarrow\OMode{#4}\dashv\OMode{#3}}
+\newcommand\NoRefine[2]{\IMode{#1}\Vdash\IMode{#2}\Uparrow}
 
 \newcommand\ThenMultiAux[5]{%
-  \InputMode{#2}\dashv\InputMode{#3}%
-  \Vdash^\star\InputMode{#1}\Rightarrow%
-  \OutputMode{#4}\dashv\OutputMode{#5}%
+  \IMode{#2}\dashv\IMode{#3}%
+  \Vdash^\star\IMode{#1}\Rightarrow%
+  \OMode{#4}\dashv\OMode{#5}%
 }
 
-\newcommand\Member[2]{\InputMode{#1}\in\InputMode{#2}}
+\newcommand\Member[2]{\IMode{#1}\in\IMode{#2}}
 \newcommand\MemberUnmoded[2]{#1\in #2}
-\newcommand\IsTrue[1]{\InputMode{#1}\ \textit{true}}
+\newcommand\EqMemberUnmoded[3]{#1 = #2\in #3}
+\newcommand\IsTrue[1]{\IMode{#1}\ \textit{true}}
 \newcommand\IsTrueUnmoded[1]{#1\ \textit{true}}
@@
 \newcommand\ThenLTac[2]{#1;[#2]}
-\newcommand\DProdIntroRule[1]{\mathtt{PairIntro}\{#1\}}
-\newcommand\DProdIntroRuleNullary{\mathtt{PairIntro}}
-\newcommand\TyProd[2]{#1 \times #2}
-\newcommand\TyDProd[3]{(#2:#1) \times #3}
+\newcommand\DPairIntroRule[1]{\mathtt{PairIntro}\{#1\}}
+\newcommand\DPairIntroRuleNullary{\mathtt{PairIntro}}
+\newcommand\DPairElimRule[1]{\mathtt{PairElim}\{#1\}}
+\newcommand\DFunIntroRule{\mathtt{FunIntro}}
+\newcommand\DFunElimRule[1]{\mathtt{FunElim}\{#1\}}
+\newcommand\TyDPair[3]{(#2:#1) \times #3}
+\newcommand\TyDFun[3]{(#2:#1) \to #3}
+\newcommand\Lam[2]{\lambda(#1.#2)}
+\newcommand\Ap[2]{\mathsf{ap}(#1;#2)}
+\newcommand\Ax{\mathsf{Ax}}
+\newcommand\Tuple[1]{\left\langle#1\right\rangle}
+\newcommand\Spread[4]{\mathsf{spread}(#1;#2,#3.#4)}
+\newcommand\TyUniv{\mathbb{U}}
 \newcommand\TyMember[2]{#1\in#2}
 \newcommand\Dom[1]{\left\vert#1\right\vert}
-\newcommand\SubsetEq[2]{\InputMode{#1}\subseteq\InputMode{#2}}
+\newcommand\SubsetEq[2]{\IMode{#1}\subseteq\IMode{#2}}
 \newcommand\FV[1]{\textit{FV}\left(#1\right)}
@@
 
-\newcommand\IsEquiv[2]{\InputMode{#1}\equiv\InputMode{#2}}
+\newcommand\IsEquiv[2]{\IMode{#1}\equiv\IMode{#2}}
+
+\newcommand\Seq[2]{#1 \gg #2}
+
+\newcommand\Ext[1]{{\color{dark-gray}\left\lfloor\mathbf{ext}\ \OMode{#1}\right\rfloor}}
+\newcommand\SeqExt[3]{
+  \Seq{\IMode{#1}}{\IMode{#2}}\ \Ext{#3}
+}
+\newcommand\NuprlRule[3]{%
+  \begin{array}{l}%
+    #2\\
+    \ \ \mathbf{by}\ #1\\
+    \quad #3
+  \end{array}%
+}
 
@@
 \[
-    \InputMode{\ThenTac{\tau_1}{\tau_2}}
+    \IMode{\ThenTac{\tau_1}{\tau_2}}
     \triangleq
-    \OutputMode{\ThenTac{\tau_1}{\MkSet{\tau_2}}}
+    \OMode{\ThenTac{\tau_1}{\MkSet{\tau_2}}}
 \]
@@
   \infer[\textit{PairIntro}]{
-    \Member{\Tuple{M,N}}{\TyDProd{A}{x}{B}}
+    \Member{\Tuple{M,N}}{\TyDPair{A}{x}{B}}
   }{
@@
 However, consider what a refinement rule for this would look like; we would
-like to take the goal $\IsTrue{\TyDProd{A}{x}{B}}$ to two subgoals (one for
+like to take the goal $\IsTrue{\TyDPair{A}{x}{B}}$ to two subgoals (one for
 each conjunct); the first subgoal $\IsTrue{A}$ is clear enough, but it is not
@@
 subgoals, we cannot give a refinement treatment to this rule, and must
-instead write a family of refinement rules $\DProdIntroRule{w}$ fibred over
+instead write a family of refinement rules $\DPairIntroRule{w}$ fibred over
 witnesses of the left conjunct $w$:
@@
 \[
-  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRule{w}}{%
+  \Refine{\IsTrueUnmoded{\TyDPair{A}{x}{B}}}{\DPairIntroRule{w}}{%
     \left\{\begin{array}{l}
@@
 We can now encode \emph{PairIntro} with a single refinement rule,
-$\DProdIntroRuleNullary$:
+$\DPairIntroRuleNullary$:
 
 \[
-  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRuleNullary}{%
+  \Refine{\IsTrueUnmoded{\TyDPair{A}{x}{B}}}{\DPairIntroRuleNullary}{%
     \left\{\begin{array}{l}
@@
     \ThenMultiAux{
-      \MkFam{\tau}{\beta}{\Dom{\ISG{\alpha}{\JJ},\Phi}}
+      \MkFam{\tau}{\beta}{\MkSet{\alpha}\cup\Dom\Phi}
     }{E}{
@@
   }{
-    \Refine{\JJ}{\tau}{\Psi_\alpha}{E_\alpha} &
+    \Refine{\JJ}{\tau_\alpha}{\Psi_\alpha}{E_\alpha} &
     \ThenMultiAux{
@@
 
-%% \section{ML Implementation of \DepLCF}
+\section{Case Study: Constructive Type Theory}
+
+The refinement logic for CTT is a sequent calculus with a single form of
+judgment, \framebox{$\Seq{\IMode{H}}{\IMode{P}}$}, which means that the
+proposition $P$ is true (and functionally so) under the assumptions in the
+telescope $H$. The synthesis of this judgment is a hypothetical witness for the
+truth or $P$ (or \emph{extract term}) with free variables in $H$.
+
+The type theory CTT is then approximated in the refinement logic by adding
+valid refinement rules for $\Seq{\IMode{H}}{\IMode{P}}$. For the sake of
+familiarity, we will re-use Nuprl's rule notation such that
+\[
+  \NuprlRule{\texttt{RuleName}}{
+    \SeqExt{H}{P}{E(\alpha_0,\alpha_1,\dots,\alpha_n)}
+  }{
+    \begin{array}{l}
+      \SeqExt{H_0}{Q_0}{\alpha_0}\\
+      \SeqExt{H_1(\alpha_0)}{Q_1(\alpha_0)}{\alpha_1(\alpha_0)}\\
+      \vdots\\
+      \SeqExt{H_n(\alpha_0,\alpha_1,\dots,\alpha_{n-1})}{Q_n(\alpha_0,\alpha_1,\dots,\alpha_{n-1})}{\alpha_n(\alpha_0,\dots,\alpha_{n-1})}
+    \end{array}
+  }
+\]
+shall mean
+\[
+  \Refine{
+    \Seq{H}{P}
+  }{
+    \mathtt{RuleName}
+  }{
+    \left\{\begin{array}{l}
+      \OSG{\alpha_0}{\Seq{H_0}{Q_0}}\\
+      \OSG{\alpha_1}{\Seq{H_1}{Q_1}}\\
+      \multicolumn{1}{c}{\color{Black}\vdots}\\
+      \OSG{\alpha_n}{\Seq{H_n}{Q_n}}\\
+    \end{array}\right.
+  }{
+    E(\alpha_0,\alpha_1,\dots,\alpha_n)
+  }
+\]
+
+Let us first give some of the refinement rules for the dependent pair type:
+
+\begin{gather*}
+  \NuprlRule{\DPairIntroRuleNullary}{
+    \SeqExt{H}{\TyDPair{A}{x}{B}}{\Tuple{\alpha,\beta}}
+  }{
+    \begin{array}{l}
+      \SeqExt{H}{A}{\alpha}\\
+      \SeqExt{H}{\Subst{\alpha}{x}{B}}{\beta}\\
+      \SeqExt{H,x:A}{\MemberUnmoded{B}{\TyUniv}}{\gamma(x)}
+    \end{array}
+  }\\[6pt]
+  \NuprlRule{\DPairElimRule{z}}{
+    \SeqExt{H,z:\TyDPair{A}{x}{B},J}{C}{
+      \Spread{z}{u}{v}{\alpha(u,v)}
+    }
+  }{
+    \SeqExt{H,z:\TyDPair{A}{x}{B},u:A,v:\Subst{u}{x}{B},\Subst{\Tuple{u,v}}{z}{J}}{C}{
+      \alpha(u,v)
+    }
+  }
+\end{gather*}
+
+We can also give refinement rules for dependent function introduction and
+elimination; in the same way \ModLCF\ could not express a refinement rule for
+pair introduction, a refinement rule for function elimination likewise only
+becomes attainable via the generalization to \DepLCF.
+
+\begin{gather*}
+  \NuprlRule{\DFunIntroRule}{
+    \SeqExt{H}{\TyDFun{A}{x}{B}}{\Lam{x}{\alpha(x)}}
+  }{
+    \begin{array}{l}
+      \SeqExt{H,x:A}{B}{\alpha(x)}\\
+      \SeqExt{H}{\MemberUnmoded{A}{\TyUniv}}{\beta}
+    \end{array}
+  }\\[6pt]
+  \NuprlRule{\DFunElimRule{z}}{
+    \SeqExt{
+      H,z:\TyDFun{A}{x}{B},J
+    }{C}{
+      \beta(\Ap{z}{\alpha}, \Ax)
+    }
+  }{
+    \begin{array}{l}
+      \SeqExt{H,z:\TyDFun{A}{x}{B}, J}{A}{\alpha}\\
+      \SeqExt{
+        H,z:\TyDFun{A}{x}{B},%
+        y:\Subst{\alpha}{x}{B},%
+        p:\EqMemberUnmoded{y}{\Ap{z}{\alpha}}{\Subst{\alpha}{x}{B}}, J
+      }{C}{
+        \beta(y,p)
+      }
+    \end{array}
+  }
+\end{gather*}
 

### 2015-10-18 doc/paper.tex
layout
wordsmithing

@@
 $\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}$ which accounts for this dependency.
-In order to do this, we will need to implement an auxiliary judgment
+To this end, we will need to implement an auxiliary judgment
 \framebox{$\ThenMultiAux{\MkFam{\tau}{\alpha}{I}}{E}{\Phi}{E'}{\Phi'}$} that
@@
 a proof state, simultaneously propagating refinements rightward through
-substitution. This judgment, which presupposes $\IsEquiv{I}{\Dom\Phi}$, is defined
-by recursion on the telescope $\Phi$, viewed as a \emph{cons}-list:
+substitution. This judgment, which presupposes $\IsEquiv{I}{\Dom\Phi}$, is
+defined by recursion on the telescope $\Phi$, viewed as a \emph{cons}-list:
 
@@
 telescope $H$. The synthesis of this judgment is a hypothetical witness for the
-truth or $P$ (or \emph{extract term}) with free variables in $H$.
+truth or $P$ (or \emph{extract term}) with free variables bound in $H$.
 
@@
 
-\begin{gather*}
+\[
+  \begin{array}{l}
     \NuprlRule{\DPairIntroRuleNullary}{
@@
       \end{array}
-  }\\[6pt]
+    }\\\\
     \NuprlRule{\DPairElimRule{z}}{
@@
     }
-\end{gather*}
+  \end{array}
+\]
 
@@
 
-\begin{gather*}
+\[
+  \begin{array}{l}
     \NuprlRule{\DFunIntroRule}{
@@
       \end{array}
-  }\\[6pt]
+    }\\\\
     \NuprlRule{\DFunElimRule{z}}{

### 2015-10-25 doc/paper.tex
Start taking advantage of ABT framework

@@
 \newcommand\MkFam[3]{\MkSet{#1_{#2}}_{#2\in #3}}
+
+\newcommand\IsAbt[5]{%
+  \IMode{#1}\triangleright%
+  \IMode{#2}\parallel%
+  \IMode{#3}\vdash%
+  \IMode{#4}:\OMode{#5}%
+}
+\newcommand\IsWf[5]{%
+  \IMode{#1}\triangleright%
+  \IMode{#2}\parallel%
+  \IMode{#3}\vdash%
+  \IMode{#4}\in_{\mathbf{wf}}\IMode{#5}%
+}
+
+\newcommand\SortJdg{\mathsf{jdg}}
+\newcommand\MkBTm[3]{\{#1\}[#2].\, #3}
+\newcommand\JdgValence[1]{\mathfrak{v}(#1)}
+\newcommand\Metavars[1]{\mathfrak{M}(#1)}
 
@@
 a \ModLCF\ refiner judgmentally via two forms of judgment,
-\framebox{$\Refine{\JJ}{\tau}{\Psi}{E}$} and \framebox{$\NoRefine{\JJ}{\tau}$}, where $\JJ$ is a judgment of the
-logical theory, $\tau$ is a tactic, $\Psi$ is a context of judgments,
-representing the subgoals generated by the tactic $\tau$, and $E$ is the
-synthesis of the judgment $\JJ$, binding variables $\Dom{\Psi}$ which represent
-the syntheses of the subgoals.
+\framebox{$\Refine{\JJ}{\tau}{\Psi}{E}$} and \framebox{$\NoRefine{\JJ}{\tau}$},
+where $\JJ$ is a judgment of the logical theory, $\tau$ is a tactic, $\Psi$ is
+a context of judgments, representing the subgoals generated by the tactic
+$\tau$, and $E$ is the synthesis of the judgment $\JJ$, binding variables
+$\Dom{\Psi}$ which represent the syntheses of the subgoals.
 
 The meaning of $\Refine{\JJ}{\tau}{\Psi}{E}$ is that $\tau$ is applicable to
-demonstrating the judgment $\JJ$, producing synthesis $e$ under the assumptions
+demonstrating the judgment $\JJ$, producing synthesis $E$ under the assumptions
 that the judgments in $\Psi$ can be demonstrated. The divergence judgment
@@
 
+In order to properly treat generic judgments whose syntheses bind variables, we
+will commit up front to using the abstract binding trees logical framework as a
+universal domain of discourse, with a distinguished sort $\SortJdg$ for
+object-judgments; and to each judgment $\JJ$ such that
+$\IsWf{\Omega}{\Upsilon}{\Gamma}{\JJ}{\SortJdg}$, a \emph{valence}
+$\JdgValence{\JJ}$ shall be associated which describes the binding structure of
+its synthesis. Then, to each subgoal context $\Psi$, we shall associate a
+metavariable context $\Metavars{\Psi}$ as follows:
+
+\begin{align*}
+  \IMode{\Metavars{\cdot}} &= \OMode{\cdot}\\
+  \IMode{\Metavars{\Psi, \alpha:\JJ}} &= \OMode{\Metavars{\Psi}, \alpha : \JdgValence{\JJ}}
+\end{align*}
+
+Then, the judgments $\NoRefine{\JJ}{\tau}$ and $\Refine{\JJ}{\tau}{\Psi}{E}$
+shall presuppose $\IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}$, and the latter
+shall postsuppose $\IsWf{\Metavars{\Psi}}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$.
+When $E$ binds no variables or symbols, we shall write $E$ instead of
+$\MkBTm{\cdot}{\cdot}{E}$.
+
+%% In practice, this means that the
+%% judgment $\Refine{\JJ}{\tau}{\Psi}{E}$ shall presuppose
+%% $\IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}$
+
 Let $\JSet$ be the open-ended collection of judgments in our logical theory,
@@
 $\Refine{\JJ}{R}{\Psi}{E}$ and $\NoRefine{\JJ}{R}$ must be explained for
-$\Member{J}{\JSet}$. We say \framebox{$\IsTac{\tau}$} in case for all object-judgments
-$\Member{\JJ}{\JSet}$, the assertion conditions for
-$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$ are disjoint, and
-moreover, if $\Refine{\JJ}{\tau}{\Psi}{E}$, then
-$\SubsetEq{\FV{E}}{\Dom{\Psi}}$.
+$\Member{J}{\JSet}$. We say \framebox{$\IsTac{\tau}$} in case for all
+object-judgments $\Member{\JJ}{\JSet}$, the assertion conditions for
+$\Refine{\JJ}{\tau}{\Psi}{E}$ and $\NoRefine{\JJ}{\tau}$ are well-defined and
+disjoint.
 
@@
 \begin{thm}
-  The above rules all define valid tactics:.
+  The above rules all wellformed definitions of tactics.
   \begin{enumerate}
@@
       $\IsTac{\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}}$.
-
-     %% If $\IsTac{\tau_1}$ and $\IsTac{\tau_2}$, then
-     %% $\IsTac{\ThenTac{\tau_1}{\tau_2}}$.
   \end{enumerate}
@@
 \begin{proof}
-  It suffices to verify that the synthesis $E$ of each tactical is well-scoped
-  in $\Psi$; in each case, this follows by induction.
+  Disjointness for each tactical is immediately evident; it suffices, then, to
+  verify the postsupposition
+  $\IsWf{\Metavars{\Psi}}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$ in each case.
   \begin{enumerate}
@@
       $\SubsetEq{\FV{E}}{\Dom\Phi}$ and that for each $\Member{\alpha}{\Phi}$,
-      $\SubsetEq{\FV{F_\alpha}}{\Dom{\Psi_\alpha}}$; all of the free variables
+      $\SubsetEq{\FV{F_\alpha}}{\Dom{\Psi_\alpha}}$; all of the free metavariables
       of the term got by substituting each $F_\alpha$ for
@@
 
+%%% TODO: correct to use new metavariable kit!!
+
 The refinement logic for CTT is a sequent calculus with a single form of

### 2015-10-25 doc/paper.tex
Correct ctx wellformedness; add generalization to telescopes

@@
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\usepackage{amsmath, amsthm, stmaryrd, scalerel}
+\usepackage{amsmath, amssymb, amsthm, stmaryrd, scalerel}
 \usepackage{mathtools}
@@
 
+\newcommand\IsCtx[2]{
+  \IMode{#1}\ \textit{ctx}\leadsto \OMode{#2}
+}
+\newcommand\IsTele[2]{
+  \IMode{#1}\ \textit{tele}\leadsto \OMode{#2}
+}
+
 \newcommand\SortJdg{\mathsf{jdg}}
@@
 \newcommand\JdgValence[1]{\mathfrak{v}(#1)}
-\newcommand\Metavars[1]{\mathfrak{M}(#1)}
 
@@
 $\JdgValence{\JJ}$ shall be associated which describes the binding structure of
-its synthesis. Then, to each subgoal context $\Psi$, we shall associate a
-metavariable context $\Metavars{\Psi}$ as follows:
+its synthesis.
 
-\begin{align*}
-  \IMode{\Metavars{\cdot}} &= \OMode{\cdot}\\
-  \IMode{\Metavars{\Psi, \alpha:\JJ}} &= \OMode{\Metavars{\Psi}, \alpha : \JdgValence{\JJ}}
-\end{align*}
+Then, we will need to define the wellformedness of subgoal contexts $\Psi$,
+simultaneously associating to $\Psi$ a metavariable context $\Omega$ which will
+stand for the collected syntheses of $\Psi$:
+\begin{gather*}
+  \infer{
+    \IsCtx{\cdot}{\cdot}
+  }{
+  }\qquad
+  \infer{
+    \IsCtx{\Psi,\alpha:\JJ}{\Omega,\alpha:\JdgValence{\JJ}}
+  }{
+    \IsCtx{\Psi}{\Omega} &
+    \IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}
+  }
+\end{gather*}
+
+\begin{remark}
+  Note how the context-wellformedness judgment artificially requires that $\JJ$
+  be wellformed in the empty metavariable context.
+\end{remark}
 
@@
 shall presuppose $\IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}$, and the latter
-shall postsuppose $\IsWf{\Metavars{\Psi}}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$.
+shall postsuppose $\IsCtx{\Psi}{\Omega}$ and $\IsWf{\Omega}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$.
 When $E$ binds no variables or symbols, we shall write $E$ instead of
 $\MkBTm{\cdot}{\cdot}{E}$.
-
-%% In practice, this means that the
-%% judgment $\Refine{\JJ}{\tau}{\Psi}{E}$ shall presuppose
-%% $\IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}$
 
@@
 \begin{thm}
-  The above rules all wellformed definitions of tactics.
+  The above rules all define valid tactics.
   \begin{enumerate}
@@
   Disjointness for each tactical is immediately evident; it suffices, then, to
-  verify the postsupposition
-  $\IsWf{\Metavars{\Psi}}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$ in each case.
+  verify the postsuppositions $\IsCtx{\Psi}{\Omega}$ and
+  $\IsWf{\Omega}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$ in each case.
   \begin{enumerate}
@@
 
-We can now encode \emph{PairIntro} with a single refinement rule,
-$\DPairIntroRuleNullary$:
+Briefly, we can adjust the postsuppositions of the refinement judgment to
+account for this generalization, by first characterizing the wellformed
+telescopes:
+\begin{gather*}
+  \infer{
+    \IsTele{\cdot}{\cdot}
+  }{
+  }\qquad
+  \infer{
+    \IsTele{\Psi,\alpha:\JJ}{\Omega,\alpha:\JdgValence{\JJ}}
+  }{
+    \IsTele{\Psi}{\Omega} &
+    \IsWf{\Omega}{\cdot}{\cdot}{\JJ}{\SortJdg}
+  }
+\end{gather*}
 
-\[
-  \Refine{\IsTrueUnmoded{\TyDPair{A}{x}{B}}}{\DPairIntroRuleNullary}{%
-    \left\{\begin{array}{l}
-        \OSG{\alpha}{\IsTrueUnmoded{A}}\\
-        \OSG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
-    \end{array}\right.
-  }{
-    \Tuple{\alpha,\beta}
-  }
-\]
+Now, $\Refine{\JJ}{\tau}{\Psi}{E}$ shall postsuppose $\IsTele{\Psi}{\Omega}$
+and $\IsWf{\Omega}{\cdot}{\cdot}{E}{\JdgValence{\JJ}}$.
 
+\subsection{Dependent Tacticals}
 All that remains is to give a new definition of the sequencing tactical

### 2015-10-25 doc/paper.tex
Concrete definition of \JSet
clarify

@@
 
-Let $\JSet$ be the open-ended collection of judgments in our logical theory,
+Let $\JSet$ be the set $\MkSet{\JJ\mid\IsWf{\cdot}{\cdot}{\cdot}{\JJ}{\SortJdg}}$,
 and let $\RSet$ be a collection of rule names. Each rule $\Member{R}{\RSet}$
@@
 a proof state, simultaneously propagating refinements rightward through
-substitution. This judgment, which presupposes $\IsEquiv{I}{\Dom\Phi}$, is
-defined by recursion on the telescope $\Phi$, viewed as a \emph{cons}-list:
+metavariable substitution. This judgment, which presupposes
+$\IsEquiv{I}{\Dom\Phi}$, is defined by recursion on the telescope $\Phi$,
+viewed as a \emph{cons}-list:
 
